---
title: "NYPD Shooting Incident Analysis"
output: pdf_document
date: "2024-06-01"
author: "Christopher Washington"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Violent crime is something that has the potential to affect each one of us every single day. It is important to analyze violent crime data in order to allocate resources to mitigate crime in every neighborhood in the world. Luckily for us, there is data publicly available for shootings in New York City provided by the New York Police Department (NYPD). The report will take a small look into this data.
## Required R Libraries
The following R libraries are required to run the R code contained in this document: 
```{r load_libs, echo=TRUE, message=FALSE}
library(tidyverse)
library(knitr)
library(lubridate)
library(reshape2)
set.seed(10)
```

## Dataset
The dataset that will be used for this report is the NYPD Shooting Incident Data (Historic) located at <https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD>. This dataset contains a comprehensive list of shootings in New York City from 2006 to the end of the previous calendar year. It includes data on the perpetrator and victim of the shooting, such as age, gender, and race. The dataset also contains data on location such as the borough that the incident occurred in, as well data about the date and time of the incident.

There is also population data used in this analysis that is obtained from 
the census.gov data located at <https://www.census.gov/quickfacts/fact/table/newyorkcountynewyork,richmondcountynewyork,kingscountynewyork,queenscountynewyork,bronxcountynewyork,newyorkcitynewyork/PST045222>

## Data Retrieval
We will begin by loading the dataset
```{r load_data, message=FALSE}
data_url = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
shootings = read_csv(data_url)
```
We are only interested in the date and borough for this analysis, so we will remove the unnecessary columns. We will also rename the columns to simpler names for later use.
```{r rename_cols, message=FALSE}
shootings = shootings %>% 
  rename(date = 'OCCUR_DATE') %>%
  rename(boro = 'BORO') %>%
  select(date, boro)
```
We will continue to preprocess our data by converting columns to the appropriate data types. Categorical data columns will become factors, and date/time columns will be converted to the appropriate date/time data type.
```{r convert_cols, message=FALSE}
shootings = shootings %>% mutate_at(
  vars(boro), 
  factor
)
shootings = shootings %>% mutate(date = mdy(date))
```
Finally, we will filter out any null and UNKNOWN records from the dataset:
```{r filter_rows, message=FALSE}
shootings = shootings %>% 
  filter(boro != "(null)") %>% 
  filter(boro != "UNKNOWN")
```
The final dataset looks like the following snapshot where each row contains the date and borough for a single shooting:
```{r show_data, echo=FALSE}
kable(head(shootings))
```

## Visualization 1: Shootings Per 1000 Residents in Each Borough
The first visualization that we will consider is the number of shootings per 1000 residents in each New York City borough. We compute the per 1000 value as follows
```{r shootings_per_1000, message=FALSE}
shootings_by_borough_per_1000 = shootings %>%
  group_by(boro) %>% 
  summarise(n = n()) %>% 
  mutate(pop_n =
    ifelse(boro == 'BRONX', 1472654, 
        ifelse(boro == 'BROOKLYN', 2736074, 
           ifelse(boro == 'MANHATTAN', 1694251, 
              ifelse(boro == 'QUEENS', 2405464, 
                 ifelse(boro == 'STATEN ISLAND', 495747, n)
             )
          )
       )
    )
  ) %>%
  mutate(per_1000 = round(1000*(n/pop_n), 2)) %>%
  select(boro, per_1000)
```
and we obtain the following data
```{r shooings_per_1000_table, echo=FALSE}
kable(shootings_by_borough_per_1000)
```
We then plot this data on a bar chart\newline
```{r shooings_per_1000_chart, echo=FALSE}
shootings_by_borough_per_1000 %>% ggplot() +
  geom_col(aes(
    x = boro,
    y = per_1000)) +
  labs(x = "Borough", y="Count") +
  ggtitle("Shootings by Borough (Per 1000 Residents)") +
  theme_bw() +
  theme(axis.text.x = element_text(face = 'bold', size = 10),
        axis.text.y = element_text(face = 'bold', size = 10))
```

## Model 1: Shootings by Borough vs Population by Borough
We will now conduct some analysis of the distribution of shootings in each borough. If there are no factors contributing to increases or decreases in shootings in each borough, then we can assume that the proportions of shootings in each borough will reflect the populations in each borough. We compute the proportions of shootings in each borough, and the proportion of population in each borough as follows
```{r compute_proportions, message=FALSE}
shootings_by_borough = shootings %>%
  group_by(boro) %>% 
  summarise(n = n()) %>% 
  mutate(p = round(n/sum(n), 4)) %>%
  mutate(pop_n =
    ifelse(boro == 'BRONX', 1472654, 
        ifelse(boro == 'BROOKLYN', 2736074, 
           ifelse(boro == 'MANHATTAN', 1694251, 
              ifelse(boro == 'QUEENS', 2405464, 
                 ifelse(boro == 'STATEN ISLAND', 495747, n)
             )
          )
       )
    )
  ) %>%
  mutate(pop_p =
    ifelse(boro == 'BRONX', round(1472654/sum(pop_n), 4), 
        ifelse(boro == 'BROOKLYN', round(2736074/sum(pop_n), 4), 
           ifelse(boro == 'MANHATTAN', round(1694251/sum(pop_n), 4), 
              ifelse(boro == 'QUEENS', round(2405464/sum(pop_n), 4), 
                 ifelse(boro == 'STATEN ISLAND', round(495747/sum(pop_n), 4), n)
             )
          )
       )
    )
  )
```
From this computation we obtain the following data
```{r proportions_table, echo=FALSE}
kable(shootings_by_borough)
```
To test if it is indeed the case that the proportion of shootings in each borough matches the proportion of population in each borough, we will use the Chi-Squared Goodness of Fit test. The test statistic we will consider is $$\sum_{i=1}^{n}\frac{(O_i-E_i)^2}{E_i}$$. We compute the test statistic as follows
```{r proportions_test_stat, message=FALSE}
shootings_by_borough_analysis = shootings_by_borough %>%
  mutate(exp_n = round(sum(n)*pop_p, 0)) %>%
  mutate(diff = n-exp_n) %>%
  mutate(diff_sq = diff^2) %>%
  mutate(diff_sq_scl = round(diff_sq/exp_n, 0)) %>%
  select(boro, n, exp_n, diff, diff_sq, diff_sq_scl)
```
We obtain the following data
```{r test_stat_table, echo=FALSE}
kable(shootings_by_borough_analysis)
```
Then we sum up the last column of the previous table to obtain the test statistic value of 5,941. The p-value of this test statistic is given by
```{r test_stat_p_value}
1-pchisq(5941, 4)
```
Since the p-value is extremely small, we may reject the assumption that the proportion of shootings in each borough matches the population in each borough. We visualize this different in the following bar chart displaying the number of shootings in each borough (n) next to the expected number of shootings in each borough under the previous assumption (exp_n).\newline
```{r build_test_borough_model, echo=FALSE, message=FALSE}
melt(shootings_by_borough_analysis %>% select(boro, n, exp_n), id = c("boro")) %>%
  ggplot(aes(
    x = boro, 
    y = value, 
    fill = variable)) +
  labs(x = "Borough", y = "Proportion") + 
  ggtitle("Proportions") +
  theme_bw() +
  geom_bar(stat="identity", width=0.8, position = "dodge")  

```

## Analysis
Both visualtions displayed here raise questions about the distributions of shootings in each borough. It is clear that some boroughs like Bronx and Brooklyn have more shooting incidents than others. Reasons for these discrepancies are numerous, and may include socioeconomic differences or geographic differences. In future work we can explore these other topics to attempt to zero in on the causes of the differences.

## Bias Identification
We do not see any bias in the dataset that we have constructed, or the analyses we have made here. The dataset simply contains a date and location for each shooting. However, it is possible that bias occurs in the underlying dataset. It may be the case that the dataset does not contain every shooting instance in New York City, but only contains the incidents that the police were called and investigated for. This could certainly upset the visualizations and analysis that we have done here.

## Conclusion 
There are many other visualization and analyses that can be brought from this dataset. Two of those have been provided here. These visualizations and analyses provide a starting point for further research and anlysis regarding shootings in New York City.